The hydra and the rooster

### 2020-07-24 latex/chapter-ordinals.tex
redaction introduction to ordinals

The following class definition, parameterized with a type $A$, a binary relation \texttt{lt} on $A$,

@@

\begin{Coqsrc}
Class OrdinalNotation {A:Type}{lt: relation A}(sto:StrictOrder lt)
      (compare : A -> A -> comparison)  :=
  {   wf : well_founded lt;
      compare_correct :
         forall alpha beta:A,
          CompareSpec (alpha=beta) (lt alpha beta) (lt beta alpha)
                                   (compare alpha beta);
  }.

Definition on_t  {A:Type}{lt: relation A}
            {compare : A -> A -> comparison}
            {on : OrdinalNotation lt compare} := A.

Definition on_compare {A:Type}{lt: relation A}
            {compare : A -> A -> comparison}
            {on : OrdinalNotation lt compare} := compare.

Definition on_lt {A:Type}{lt: relation A}
           {compare : A -> A -> comparison}
           {on : OrdinalNotation lt compare} := lt.

Definition on_le  {A:Type}{lt: relation A}
           {compare : A -> A -> comparison}
           {on : OrdinalNotation lt compare} :=
  clos_refl _ on_lt.

>>>>

The following class definition, parameterized with a type $A$, a binary relation \texttt{lt} on $A$, specifies that \texttt{lt} is a well-founded strict order. The reflexive closure of \texttt{lt}, (called \texttt{le}, for ``less  or equal than'') is a total decidable order, implemented through a comparison function \texttt{compare}.

@@

\begin{Coqsrc}
Class ON {A:Type}(lt: relation A)
      (compare : A -> A -> comparison)  :=
  {
  sto :> StrictOrder lt;
  wf : well_founded lt;
  compare_correct :
    forall alpha beta:A,
      CompareSpec (alpha=beta) (lt alpha beta) (lt beta alpha)
                  (compare alpha beta);
  }.
\end{Coqsrc}

The following definitions allow us to make implicit several guessable arguments.

\begin{Coqsrc}
Definition on_t  {A:Type}{lt: relation A}
            {compare : A -> A -> comparison}
            {on : ON lt compare} := A.

Definition on_compare {A:Type}{lt: relation A}
            {compare : A -> A -> comparison}
            {on : ON lt compare} := compare.

Definition on_lt {A:Type}{lt: relation A}
           {compare : A -> A -> comparison}
           {on : ON lt compare} := lt.
Infix "<" := on_lt : ON_scope.

Definition on_le  {A:Type}{lt: relation A}
           {compare : A -> A -> comparison}
           {on : ON lt compare} :=
  clos_refl _ on_lt.

Infix "<=" := on_le : ON_scope.
\end{Coqsrc}

\subsection{Comparing Ordinal Notations}

It is sometimes useful to compare two ordinal notations with respect to expressive power
(the segment of ordinals one they represent).

The following class specifies a strict inclusion of segments. The notation \texttt{OA} describes a segment $[0,alpha($, and \texttt{OB} is a larger segment (which contains a notation for $\alpha$). We require that the comparison functions of the two notation systems are compatible.

\begin{Coqsrc}
Class  SubON 
       `(OA : @ON A ltA  compareA)
       `(OB : @ON B ltB  compareB)
       (alpha :  B)
       (iota : A -> B):=
  {
  subseg_compare :forall x y : A,  compareB (iota x) (iota y) =
                                 compareA x y;
  subseg_incl : forall x, ltB (iota x) alpha;
  subseg_onto : forall y, ltB y alpha  -> exists x:A, iota x = y}.
\end{Coqsrc}

\begin{todo}
pointers to examples 
\end{todo}

In some cases we want to show that two notation systems describe the same segment (for instance $3+\omega$ and $\omega$). For that purpose, one may prove that the two notation systems are order-isomorphic.

\begin{Coqsrc}
Class  ON_Isomorphic 
       `(OA : @ON A ltA compareA)
       `(OB : @ON B ltB  compareB)
       (f : A -> B)
       (g : B -> A):=
  {
  iso_compare_comm :forall x y : A,  compareB (f x) (f y) =
                                 compareA x y;
  iso_inv1 : forall a, g (f a)= a;
  iso_inv2 : forall b, f (g b) = b}.
\end{Coqsrc}

Finally, it may be interesting to compare an ordinal notation with the more theoretical model from Schütte (well, at least with our formalization of thhat model). This would be a relative proof of correctenss of our ordinal  notations.

The following class specifies that a notation \texttt{OA} describes a segment $[0,\alpha)$,
where $\alpha$ is a countable ordinal à la Schütte.

\begin{Coqsrc}
Class ON_for `(alpha : Schutte_basics.Ord)
     `(OA : @ON A ltA  compareA)
      (iota : A -> Schutte_basics.Ord) :=
  { conform_inj : forall a, Schutte_basics.lt (iota a) alpha;
    conform_surj : forall beta, Schutte_basics.lt beta alpha ->
                                exists b, iota b = beta;
    conform_compare : forall a b:A,
        match compareA a b with
          Datatypes.Lt => Schutte_basics.lt (iota a) (iota b)
        | Datatypes.Eq => iota a = iota b
        | Datatypes.Gt => Schutte_basics.lt (iota b) (iota a)
        end}.
\end{Coqsrc}

For instance, the following theorem tells that \texttt{Epsilon0}, our otation system for the segment $[0,\epsilon0)$ is correct with respect to the countqable ordinal $\epsilon_0$
(see chapter~\ref{chap:schutte} for more details).

\begin{Coqsrc}
Instance Epsilon0_correct :
  ON_for epsilon0 Epsilon0  (fun alpha => inject (cnf alpha)).
\end{Coqsrc}

\subsection{Examples}

The simplest example of ordinal notation is
